Simulation Learning in Computer Hex

### 2011-10-17 thesis/thesis.tex
Improved results tables.

\documentclass[11pt]{report}
\usepackage[latin1]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{url}
\newcommand{\black}{\texttt{Black}}
\newcommand{\white}{\texttt{White}}
\newcommand{\blacks}{\texttt{Black's}}
\newcommand{\whites}{\texttt{White's}}
\newcommand{\loc}[1]{\texttt{#1}}
\begin{document}

@@

This thesis presents a slightly different approach using evolutionary learning. It solves the problem of requiring a strong outside player or expert knowledge by purely using games of self-play.
% TODO: rephrase, introduce Hivemind

@@

\section{State of the Art in Monte Carlo Tree Search}\label{mcts}
As stated, the current top computer Hex player, MoHex, uses the open-source Fuego implementation of Monte-Carlo Tree Search (MCTS). The basic Monte-Carlo technique for Computer Go was proposed by Bru\"{u}gmann\cite{brugmann1993monte}. His program \emph{Mango} effectively used a ``flat'' search tree (1-ply) and simply ran simulations from each candidate move to estimate its value. To contrast, MCTS uses Monte-Carlo simulations to grow a minimax search tree. The UCT algorithm was developed alongside MCTS to direct the growth of the search tree\cite{gelly2006exploration}. At any point, the search can be stopped and the current state of the tree used to select a move to play. Performance of the search is primarily based on three factors: the \emph{selection policy}, the \emph{simulation policy}, and the \emph{update policy}.
% TODO: enumerate and define 3 different policy types. Possibly with graphic.

@@

\subsection{Simulation Policy}
The \texttt{simulate} function plays out the rest of the game from the current position. The simplest policy plays a legal uniform random move until the game ends (in Hex, when one player has a winning path). The simulation policy has the largest effect on the strength of the entire search. %TODO: justify
Initial research in Computer Go focused on using a strong simulation policy, using expert-derived heuristics to select the next move. The top Computer Go programs all have complex hand-crafted policies with rules to examine the current position and select a ``good'' move to play, possibly with a stochastic choice between several candidates. Only when none of the rules match do they fall back to a uniform random move. It was quickly found that building a stronger simulation policy does not guarantee better performance when used in the full tree search. A test and check methodology is required, with changes to the simulation policy having unpredictable effects on overall performance.

@@

\section{Problem Statement}\label{problem}
MoGo, and Go programs in general, have the benefit of a rich history of play. Thousands of expert games are available for analysis. Some approaches to improving the simulation policy have focused on using expert games in training weights to direct the playout\cite{chaslot2010adding}. More generally, common patterns are used in a hand-crafted policy to select ``urgent'' moves to play in a local area around the last played move. Hex does have one simple pattern in the form of a bridge\cite{anshelevich2002hierarchical}, but few easy paths to crafting a better simulation policy.

@@

\section{Solution}\label{solution}
% TODO: We consider simulation policies of the following forms: default, uniform local, uniform local (tenuki), learned local
After each move, a random move is selected in the 6-connected local area. This move is used as the next move in the playout. All subsequent moves will also have their local area examined for a random response. If all 6 connected vertices are non-empty, a move is played using the uniform random policy (the \emph{default} policy). In the uniform local variation, a response is chosen uniformly from any empty moves in the local area. The uniform local variation with tenuki (play away) is similar, but there is a probability $p=\frac{1}{6}$ of using the default policy. The learned local variation assigns a weight to each empty local vertex based on the local area. These weights are used to stochastically bias the local response. The weights are trained over time through self-play using Evolution Strategies.

@@

\begin{figure}
	\begin{center}
	\includegraphics[width=300px]{graphics/local-pattern.png}
	\end{center}
	\caption{A 5x5 board with the 3 patterns shown around \blacks\ last move. Example weights are shown in the center vertex.}
	\label{fig:localpattern}
\end{figure}

\begin{figure}
	\begin{center}
	\includegraphics[width=300px]{graphics/weight-pattern-map.png}
	\end{center}
	\caption{The encoding from pattern A to a 32-bit integer. Using the weight from figure \ref{fig:localpattern}, the real value 6 would be stored at position 102 in the weight map.}
	\label{fig:encoding}
\end{figure}

@@

\subsection{The Hivemind System}

@@

\begin{itemize}
\item uniform local vs. pure random: 65.00\%
\item uniform local with tenuki vs. pure random: 58.50\%
\item learned local vs. pure random: 81.50\%
\item uniform local with tenuki vs. uniform local: 45.50\%
\item learned local vs. uniform local: 76.50\%
\item learned local vs. uniform local with tenuki: 79.00\%
\end{itemize}
% TODO: tabular form

@@

As an outside control, MoHex from the University of Alberta was compiled and ran with the default settings\cite{mohex}. Again, 200 games were played. As MoHex is a particularly strong competitor, Hivemind used 100,000 simulations per move. Default time settings for MoHex are 10 seconds, but MoHex uses some time to perform game theoretic search and pruning before the Monte-Carlo phase, and will stop the search early if it believes no better moves will be found. The pre-search can also discover ``must-play'' moves that will be immediately played in lieu of the MCTS. As a result, the time controls are not directly comparable. However, CPU time was tracked and can be used as another performance comparison.

\begin{itemize}
\item learned local vs. MoHex: 6.5\%, learned local CPU = 6.64 x MoHex
\item pure random vs. MoHex: 0\% pure random CPU = 5.55 x MoHex
\item learned local vs. pure random 92.5\% learned local CPU = 1.31 x pure random
\end{itemize}

Learned local patterns provided the only wins against MoHex, with 13 wins out of 200 played. Despite the low performance, MoHex used about 1/6th to CPU time, showing a large distance to cover to get Hivemind up to competition strength.

>>>>

\documentclass[11pt]{report}
\usepackage[latin1]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{url}
\usepackage{havannah}
\newcommand{\black}{\texttt{Black}}
\newcommand{\white}{\texttt{White}}
\newcommand{\blacks}{\texttt{Black's}}
\newcommand{\whites}{\texttt{White's}}
\newcommand{\loc}[1]{\texttt{#1}}
\begin{document}

@@

The Hivemind system uses a slightly different approach. It directs a probabilistic (Monte Carlo) tree search with learned heuristics. The problem of requiring a strong outside player or expert knowledge is solved using evolutionary learning from games of self-play. Hivemind provides a framework for Monte Carlo Tree Search, with implementations that can play Hex and Go.

@@

\section{State of the Art in Monte Carlo Tree Search}\label{mcts}

\begin{figure}
	\begin{center}
	\begin{HexBoard}[board size=5]
		\HGame{a3,c2}
	\end{HexBoard}
	\includegraphics[width=300px]{graphics/tree.png}
	\end{center}
	\caption{Partial MC Search Tree}
	\label{fig:tree}
\end{figure}

As stated, the current top computer Hex player, MoHex, uses the open-source Fuego implementation of Monte-Carlo Tree Search (MCTS). The basic Monte-Carlo technique for Computer Go was proposed by Bru\"{u}gmann\cite{brugmann1993monte}. His program \emph{Mango} effectively used a ``flat'' search tree (1-ply) and simply ran simulations from each candidate move to estimate its value. To contrast, MCTS uses Monte-Carlo simulations to grow a minimax search tree. The UCT algorithm was developed alongside MCTS to direct the growth of the search tree\cite{gelly2006exploration}. At any point, the search can be stopped and the current state of the tree used to select a move to play. Performance of the search is primarily based on three algorithms.
\begin{itemize}
\item{The \emph{selection} policy recursively traverses the tree, selecting good-looking moves for exploration.}
\item{The \emph{simulation} policy randomly or semi-randomly ``plays out'' the game from the current position until a final position.}
\item{The \emph{update} policy modifies the win/loss statistics of nodes based on the simulation results.}
\end{itemize}
Figure \ref{fig:tree} shows a partial Monte-Carlo search tree for the given position. In the example, the selection policy visited the node for \texttt{Black E4}, then \texttt{White A1}. The simulation policy ran, playing out the board with \texttt{E4} occupied by \black\ and \texttt{A1} occupied by \white. This resulted in a loss for \black. The update policy choose to update all nodes along the visited path, incrementing the wins for \white\ moves and incrementing the losses for \black\ moves.

@@

\subsection{Simulation Policy}
The \texttt{simulate} function plays out the rest of the game from the current position. The simplest policy plays a legal uniform random move until the game ends (in Hex, when one player has a winning path). The simulation policy has a large effect on the strength of the entire search and, for Go, has a large body of work dedicated to improving it.
Initial research in Computer Go focused on using a strong simulation policy, using expert-derived heuristics to select the next move\cite{chaslot2010adding}. The top Computer Go programs all have complex hand-crafted policies with rules to examine the current position and select a ``good'' move to play, possibly with a stochastic choice between several candidates. Only when none of the rules match do they fall back to a uniform random move. It was quickly found that building a stronger simulation policy does not guarantee better performance when used in the full tree search\cite{gelly2006modification}. A test and check methodology is required, with changes to the simulation policy having unpredictable effects on overall performance.

@@

\section{Problem Statement}\label{problem}
MoGo specifically, and Go programs in general, have the benefit of a rich history of play. Thousands of expert games are available for analysis. Some approaches to improving the simulation policy have focused on using expert games in training weights to direct the playout\cite{chaslot2010adding}. More generally, common patterns are used in a hand-crafted policy to select ``urgent'' moves to play in a local area around the last played move. Hex does have one simple pattern in the form of a bridge\cite{anshelevich2002hierarchical}, but few easy paths to crafting a better simulation policy.

@@

\section{Solution}\label{solution}
The simulation policy has the task of quickly selecting a legal move from an arbitrary position. The policy is applied iteratively until the game is finished. Ideally, several thousand games are simulated each second, so the simulation policy must be relatively simple and fast, compared to the full tree search. We consider simulation policies of the following forms:
\begin{itemize}
	\item{Default} - The next move is chosen uniformly from the set of legal moves.
	\item{Uniform local} - The 6-connected (``local'') points surrounding the previous move are examined. The next move is chosen uniformly from any legal moves in this set. If no legal move exists on these 6 points, the default policy is used.
	\item{Uniform local with ``tenuki''} - The tenuki (Japanese: ``play away'') variant includes a small probability ($p=\frac{1}{6}$) of using the default policy, even if a legal local move exists.
	\item{Learned local} - Each local point is matched against a pattern database and assigned a weight. The weight is used as the probability of playing on the given point. Weights are trained over time through self-play using Evolution Strategies.
\end{itemize}

When used at part of a full Monte-Carlo Tree Search, the learned local policy consistently beats both the default policy and the non-learned, uniform local policies. It carries a small cost of computing the pattern hash and looking up weights in the pattern database, but most of the computation is done with offline learning. It is self-bootstrapping, using Evolution Strategies and self-play instead of a strong outside tutor or expert knowledge.

@@

\begin{figure}
	\begin{center}
	\includegraphics[width=300px]{graphics/local-pattern.png}
	\caption{A 5x5 board with the 3 patterns shown around \blacks\ last move. Example weights are shown in the center vertex.}
	\label{fig:localpattern}
	\end{center}
\end{figure}

\begin{figure}
	\begin{center}
	\includegraphics[width=300px]{graphics/weight-pattern-map.png}
	\caption{The encoding from pattern A to a 32-bit integer. Using the weight from figure \ref{fig:localpattern}, the real value 6 would be stored at position 102 in the weight map.}
	\label{fig:encoding}
	\end{center}
\end{figure}

@@

\subsection{The Hivemind System}
% TODO: Hivemind system

@@

\begin{figure}
	\begin{center}
		\begin{tabular}{c | c c c}
		 & default & uniform & uniform (tenuki) \\
		\hline
		uniform & 65\% & & \\
		uniform (tenuki) & 58.5\% & 45.5\% & \\
		learned & 81.5\% & 76.5\% & 79\% \\
		\end{tabular}
	\end{center}
	\caption{All-play-all tournament of the 4 Hivemind variants.}
	\label{fig:results}
\end{figure}

@@

As an outside control, MoHex from the University of Alberta was compiled and ran with the default settings\cite{mohex}. Again, 200 games were played. As MoHex is a particularly strong competitor, Hivemind used 100,000 simulations per move. Default time settings for MoHex are 10 seconds, but MoHex uses some time to perform game theoretic search and pruning before the Monte-Carlo phase, and will stop the search early if it believes no better moves will be found. The pre-search can also discover ``must-play'' moves that will be immediately played in lieu of the MCTS. As a result, the time controls are not directly comparable. However, CPU time was tracked and can be used for a cost comparison.

\begin{figure}
	\begin{center}
		\begin{tabular}{c c c c}
		Player 1 & Player 2 & Win-rate & Relative CPU \\
		\hline
		Hivemind Default & MoHex & 6.5\% & 664\% \\
		Hivemind Learned & MoHex & 0\% & 555\% \\
		Hivemind Learned & Hivemind Default & 92.5\% & 131\% \\
		\end{tabular}
	\end{center}
	\caption{Win-rate and relative CPU used of Hivemind versus MoHex. Relative CPU is the ratio of average CPU time per move for Player 1 vs. Player2.}
	\label{fig:relativecpu}
\end{figure}

Learned local patterns provided the only wins against MoHex, with 13 wins out of 200 played. Despite the low performance, MoHex used about 1/6th to CPU time, showing a large distance to cover to get Hivemind up to competition strength. In this series, Hivemind used 100,000 playouts per move, approximately 10 times more than the self-play results. This resulted in an increase in self-play strength, winning 92.5\% of the time when using learned patterns, up from 81.5\% with 10,000 moves. The relative cost for this performance increase was very reasonable, with pattern computation requiring a 31\% overhead.
